A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-05 analysis/paper/paper_latex.Rmd
update paper

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
bibliography: references.bib
csl: nature-biotechnology.csl
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage[normalem]{ulem}
  - \usepackage{lscape}
  - \usepackage{pbox}
  - \usepackage[labelformat=empty]{caption}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true}
  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
  - \newcommand{\hideFromPandoc}[1]{#1}
  - \hideFromPandoc{\let\Begin\begin}
  - \hideFromPandoc{\let\End\end}
  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
mainfont: Open Sans

@@

In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data. We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

@@

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking. This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about which method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("ti_evaluation_overview_v3.tmp.pdf", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

@@

<!-- Further considerations -->
When choosing a method, it is important to take two further points into account. First, it is important that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods. This is to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type. Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology. When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system. When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.

<!-- FOR METHOD DEVELOPERS -->
<!-- New tools: challenges and possibilities -->
Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories. Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories. Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods. Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers. To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies"), 
  15, 
  5
)
```

>>>>

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
bibliography: references.bib
csl: nature-biotechnology.csl
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage[normalem]{ulem}
  - \usepackage{lscape}
  - \usepackage{pbox}
  - \usepackage[labelformat=empty]{caption}
  - \usepackage{hyperref}
  - \definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
  - \definecolor{cerulean}{rgb}{0.0, 0.48, 0.65}
  - \definecolor{burntorange}{rgb}{0.8, 0.33, 0.0}
  - \hypersetup{colorlinks=true,linkcolor=airforceblue,citecolor=cerulean,urlcolor=burntorange}
  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
  - \newcommand{\hideFromPandoc}[1]{#1}
  - \hideFromPandoc{\let\Begin\begin}
  - \hideFromPandoc{\let\End\end}
  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
mainfont: Open Sans

@@

In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice of method depends on the trajectory topology present in the data. We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

@@

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking. This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea about which method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("overview_evaluation_v4.tmp.pdf", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Trajectories are automatically classified into one of nine trajectory types. d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

@@

<!-- Further considerations -->
When choosing a method, it is important to take two further points into account. First, it is important that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods. This is to make sure the prediction is not biased due to the given parameter setting or the particular algorithms underlying a TI method. Second, even if the expected topology is known, it can be beneficial to also try out methods which make less assumptions about the trajectory topology. When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system. When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.

<!-- A common trajectory analysis framework -->
Critical to the broad application of TI methods is that both the user's input and the method's output is standardised, so that users can easily and quickly try out different methods, compare their output, and run common analysis pipelines such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018]. A common framework to analyse trajectories is especially important when the output of methods contradict, as they often do. We illustrate this in  `r ref("fig", "example_predictions")`. Here, we predicted trajectories on a datasets containing fibroblasts which were reprogrammed towards neurons and myocytes, a bifurcating trajectory [@treutlein_dissectingdirectreprogramming_2016]. The topology of all five methods differs greatly, with Slingshot underestimating the complexity of the topology, while AGA pseudotime and Monocle DDRTree overestimated the complexity, at least compared to what was expected in the data.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("bifurcating_example.tmp.pdf", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "An example gold standard and prediction.", 
  caption_text = "As an illustration to our practical guidelines"
)
```

<!-- FOR METHOD DEVELOPERS -->
<!-- New tools: challenges and possibilities -->
Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories. Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories. Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods. Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers. To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies"), 
  15, 
  5
)
```
